---
title: "Model Comparison - R² and RMSE Plots"
output: html_document
---

This notebook loads per-run statistics from the three main 100-run random forest models (climate, phenology, and combined) and produces point-range plots comparing their performance. Mean R² and RMSE with SD and 95% confidence intervals are computed across the 100 runs for each model and summarised in a table. R² and RMSE are visualised as separate point-range plots.

The residual model is excluded from all comparisons as its response variable (`residualEcM`) is on a different scale to the other three models, making direct comparison of R² and RMSE uninformative. Its performance statistics are summarised in a separate table.

---

## Load the packages required for data handling and plotting.

```{r}
library(data.table)
library(ggplot2)

options(scipen = 999) # suppress scientific notation in printed output
```

---

## Read in per-run statistics from all four models, assign model labels, and compute RMSE.

```{r}
pull_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
save_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/ModelComparison'

# Read per-run stats from each model
climate_stats  <- fread(file.path(pull_dir, "AllClimateModelStats.csv"))
pheno_stats    <- fread(file.path(pull_dir, "AllPhenoModelStats.csv"))
combined_stats <- fread(file.path(pull_dir, "AllCombinedModelStats.csv"))
residual_stats <- fread(file.path(pull_dir, "AllResidualModelStats.csv"))

# Assign model labels and compute RMSE
climate_stats[,  `:=`(Model = "Climate",   RMSE = sqrt(prediction_error))]
pheno_stats[,    `:=`(Model = "Phenology", RMSE = sqrt(prediction_error))]
combined_stats[, `:=`(Model = "Combined",  RMSE = sqrt(prediction_error))]
residual_stats[, `:=`(Model = "Residual",  RMSE = sqrt(prediction_error))]
```

---

## Combine stats and summarise mean, SD, and 95% CI across the 100 runs per model.

```{r}
all_stats <- rbindlist(list(climate_stats, pheno_stats, combined_stats, residual_stats),
                       fill = TRUE)
all_stats[, Model := factor(Model, levels = c("Phenology", "Climate", "Combined", "Residual"))]

summary_stats <- all_stats[, .(
  N         = .N,
  Mean_R2   = round(mean(r_squared), 5),
  SD_R2     = round(sd(r_squared),   5),
  CI95_R2   = round(qt(0.975, df = .N - 1) * sd(r_squared) / sqrt(.N), 5),
  Mean_RMSE = round(mean(RMSE), 5),
  SD_RMSE   = round(sd(RMSE),   5),
  CI95_RMSE = round(qt(0.975, df = .N - 1) * sd(RMSE) / sqrt(.N), 5)
), by = Model]

summary_stats[, Model := factor(Model, levels = c("Phenology", "Climate", "Combined", "Residual"))]
summary_stats <- summary_stats[order(Model)]
```

---

## Split into main model table (climate, phenology, combined) and residual table.

```{r}
main_summary     <- summary_stats[Model != "Residual"]
main_summary[,   Model := factor(Model, levels = c("Phenology", "Climate", "Combined"))]
residual_summary <- summary_stats[Model == "Residual"]

cat("=== Main Model Comparison ===\n")
print(main_summary)

cat("\n=== Residual Model Summary ===\n")
print(residual_summary)
```

---

## Export both summary tables as CSVs.

```{r}
fwrite(main_summary,     file.path(save_dir, "ModelComparison_Summary.csv"))
fwrite(residual_summary, file.path(save_dir, "ResidualModel_Summary.csv"))
```

---

## Plot mean R² ± 95% CI for the three main models. Mean R² values are labelled above each point.

```{r}
r2_plot <- ggplot(main_summary, aes(x = Model, y = Mean_R2, colour = Model)) +
  geom_point(size = 3, shape = 21, fill = "white") +
  geom_errorbar(aes(ymin = Mean_R2 - CI95_R2, ymax = Mean_R2 + CI95_R2),
                width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = round(Mean_R2, 3)), vjust = -1, size = 3.5, colour = "black") +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Model", y = "R²") +
  theme_minimal() +
  theme(
    axis.line         = element_line(colour = "black", linewidth = 0.4),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(3, "pt")
  )

print(r2_plot)
```

---

## Export the R² comparison plot as a PNG.

```{r}
ggsave(file.path(save_dir, "ModelComparison_R2_Plot.png"),
       plot   = r2_plot,
       width  = 6,
       height = 5)
```

---

## Plot mean RMSE ± 95% CI for the three main models. Mean RMSE values are labelled above each point.

```{r}
rmse_plot <- ggplot(main_summary, aes(x = Model, y = Mean_RMSE, colour = Model)) +
  geom_point(size = 3, shape = 23, fill = "white") +
  geom_errorbar(aes(ymin = Mean_RMSE - CI95_RMSE, ymax = Mean_RMSE + CI95_RMSE),
                width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = round(Mean_RMSE, 3)), vjust = -1, size = 3.5, colour = "black") +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  coord_cartesian(ylim = c(0.0, 0.3)) +
  labs(x = "Model", y = "RMSE") +
  theme_minimal() +
  theme(
    axis.line         = element_line(colour = "black", linewidth = 0.4),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(3, "pt")
  )

print(rmse_plot)
```

---

## Export the RMSE comparison plot as a PNG.

```{r}
ggsave(file.path(save_dir, "ModelComparison_RMSE_Plot.png"),
       plot   = rmse_plot,
       width  = 6,
       height = 5)
```
