---
title: "Creating a Summary Dataframe"
output: html_document
---

This notebook summarises the per-year phenology columns in the combined dataset into per-metric mean and standard deviation columns, reducing the 190 phenology variables (19 metrics × 10 years, 2013–2022) down to 38 summary columns (19 metrics × mean + SD). The input is the combined dataframe produced by the *Creating a Combined Dataframe* notebook.

The 19 WorldClim bioclimate variables and cell identifier columns are carried over from the combined dataframe unchanged. The final table is column-ordered as cell identifiers → phenology summaries → climate variables, and exported as a CSV for use in downstream analyses.

---

## Load the packages required for data manipulation.

```{r}
library(data.table)
library(dplyr)
```

---

## Read in the raw combined dataframe produced by the *Creating a Combined Dataframe* notebook.

```{r}
combined_df <- fread('/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Raw_Combined_Data.csv')
```

---

## Identify phenology columns and initialise the summary dataframe with all non-phenology columns as a base.

```{r}
pheno_cols  <- grep("\\s\\d{4}$", names(combined_df), value = TRUE)
non_pheno   <- setdiff(names(combined_df), pheno_cols)
meansd_df   <- combined_df[, ..non_pheno]
```

---

## Compute row-wise mean and SD across the 10 yearly observations for each phenology metric, then reorder columns to: cell identifiers → phenology summaries → BIO climate variables.

```{r}
# Remove year suffix (e.g. " 2013") to define metric groups
metric_names <- sub("\\s\\d{4}$", "", pheno_cols)

# Split phenology columns into groups by metric
groups <- split(pheno_cols, metric_names)

# Loop through each metric group and compute row-wise mean and SD across years
for (metric in names(groups)) {
  cols <- groups[[metric]]
  meansd_df[[paste0(metric, " Mean")]] <- rowMeans(combined_df[, ..cols], na.rm = TRUE)
  meansd_df[[paste0(metric, " SD")]]   <- apply(combined_df[, ..cols], 1, sd, na.rm = TRUE)
}

# Reorder columns: cell, n_stems, propEcM, pheno summary cols, BIO cols
base_cols  <- c("cell", "n_stems_total", "propEcM")
bio_cols   <- grep("^BIO_", names(meansd_df), value = TRUE)
new_pheno  <- setdiff(names(meansd_df), c(base_cols, bio_cols))

setcolorder(meansd_df, c(base_cols, new_pheno, bio_cols))

dim(meansd_df)
head(meansd_df)
```

---

## Export the summary dataframe as a CSV for use in downstream analyses.

```{r}
fwrite(meansd_df, '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Summary_Combined_Data.csv')
```
