---
title: "Creating propECM Dataframe"
output: html_document
---

This notebook processes the GFBI (Global Forest Biodiversity Initiative) North American mycorrhizal dataset to generate a per-grid-cell ectomycorrhizal proportion metric (`propEcM`). The raw dataset contains stem counts disaggregated by mycorrhizal association type — including arbuscular mycorrhizal (AM), ectomycorrhizal (EcM), and mixed EcM-AM associations — for each GFBI grid cell across North American forest inventory plots.

For each grid cell, EcM and mixed EcM-AM stem counts are combined and divided by the total stem count to produce a proportional measure of ectomycorrhizal dominance. The output is a three-column summary table containing the grid cell identifier (`cell`), total stem count (`n_stems_total`), and the ectomycorrhizal proportion (`propEcM`). This table is exported as a CSV for use as the base dataframe in downstream analyses.

---

## Load the packages required for data import and manipulation.

```{r}
library(data.table)
library(dplyr)
```

---

## Define the file path to the GFBI North American mycorrhizal dataset. This file contains per-cell stem counts broken down by mycorrhizal association type across North American forest inventory plots.

```{r}
myc_file <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/North_America_Myc_XY_Data.csv'
```

---

## Read the mycorrhizal dataset into memory as a data table.

```{r}
myc <- fread(myc_file)
```

---

## Calculate `propEcM` for each grid cell and join the total stem count onto the result. Both pure EcM stems and mixed EcM-AM stems are included in the numerator.

```{r}
# Calculate propEcM per grid cell
ECMdf <- myc[, .(
  propEcM = (EcM + `EcM-AM`) / n_stems_total
), by = cell]

# Extract n_stems_total from the raw data and join onto ECMdf
# n_stems_total is not carried through the aggregation above, so it is pulled separately and merged back
stems <- myc[, .(cell, n_stems_total)]
ECMdf <- stems[ECMdf, on = 'cell']

print(ECMdf)
```

---

## Export the processed dataframe as a CSV. The output contains three columns: `cell` (GFBI grid cell identifier), `n_stems_total` (total stem count per cell), and `propEcM` (proportion of stems that are ectomycorrhizal).

```{r}
fwrite(ECMdf, '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Simple_Myc_Data.csv')
```

```{r}
sessionInfo()
```
```{r}
citation()
citation("data.table")
citation("dplyr")
```

