---
title: "Creating a Combined Dataframe"
output: html_document
---

This notebook merges per-cell ectomycorrhizal proportion data with remotely sensed phenology and bioclimate variables into a single wide-format table for use in downstream modelling. The ectomycorrhizal proportion data (`propEcM`) is produced by the *Creating propECM Dataframe* notebook and serves as the base table onto which all additional variables are joined by grid cell identifier (`cell`).

Phenological data are sourced from the VIIRS VNP22Q2 land surface phenology product, extracted per GFBI grid cell via Google Earth Engine across ten years of observations. Climate data are drawn from the WorldClim bioclimate variable dataset (19 bioclimate variables), also extracted via Google Earth Engine. After merging, phenology fill values are recoded to NA, timing metrics are corrected with a year-specific day-of-year offset, and temperature bioclimate variables are rescaled from their raw integer encoding to degrees Celsius.

All climate and phenology columns are renamed to human-readable labels before the combined table is exported as a CSV.

---

## Load the packages required for data import and manipulation.

```{r}
library(data.table)
library(dplyr)
```

---

## Define file paths for all input data sources.

```{r}
propECM_file <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Simple_Myc_Data.csv'
clim_folder <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Google Earth Engine/WorldClimData'
pheno_folder <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Google Earth Engine/PhenologyData'
```

---

## Read in the pre-computed ectomycorrhizal proportion data. This serves as the base table — all phenology and climate data will be merged onto it by `cell`.

```{r}
combined_df <- fread(file = propECM_file)
combined_df[, cell := as.integer(cell)]
setkey(combined_df, cell)
```

---

## Define a helper function to read and standardise a single phenology CSV from GEE. The band name is taken from the filename and used as the column name. `cell` is set as the key for merging.

```{r}
read_one_pheno <- function(f) {
  dt <- fread(f)
  dt[, cell  := as.integer(cell)]
  dt[, value := as.numeric(value)]
  dt <- dt[, .(cell, value)]
  new_name <- tools::file_path_sans_ext(basename(f))
  setnames(dt, "value", new_name)
  setkey(dt, cell)
  dt
}
```

---

## Read and merge all 190 phenology CSVs (19 bands × 10 years) into a single wide-format table, joining on `cell`.

```{r}
pheno_files <- list.files(pheno_folder, pattern = "\\.csv$", full.names = TRUE)

for (f in pheno_files) {
  ph <- read_one_pheno(f)
  combined_df <- merge(combined_df, ph, by = "cell", all.x = TRUE)
}

dim(combined_df)
```

---

## Recode phenology fill values to NA. Per the VNP22Q2 user guide, 16-bit phenology layers use 32767 and 8-bit quality layers use 255 as their no-data fill values.

```{r}
pheno_cols <- grep("^gfbi_vnp22q2_", names(combined_df), value = TRUE)

# Identify 16-bit phenology columns (fill = 32767) and 8-bit quality columns (fill = 255)
cols_32767 <- grep(
  "(Onset_Greenness_(Increase|Maximum|Decrease|Minimum)|Date_Mid_(Greenup|Senescence)_Phase|Growing_Season_Length|EVI2_|Rate_Greenness_|EVI2_Growing_Season_Area)",
  pheno_cols, value = TRUE
)
cols_32767 <- cols_32767[!grepl("PGQ", cols_32767)]

cols_255 <- grep(
  "(GLSP_QC|PGQ_|Greenness_Agreement_Growing_Season)",
  pheno_cols, value = TRUE
)

combined_df[, (cols_32767) := lapply(.SD, function(x) { x[x == 32767] <- NA; x }), .SDcols = cols_32767]
combined_df[, (cols_255)   := lapply(.SD, function(x) { x[x == 255]   <- NA; x }), .SDcols = cols_255]
```

---

## Apply a year-specific day-of-year offset to the six phenological timing metrics. Raw VNP22Q2 timing values can exceed 366, indicating a cumulative encoding relative to the year 2000. The year is extracted from each column name and used to compute the appropriate offset.

```{r}
# Select the six timing metric columns, excluding quality flag columns
timing_cols <- grep(
  "(^|_)Onset_Greenness_(Increase|Maximum|Decrease|Minimum)_|(^|_)Date_Mid_(Greenup|Senescence)_Phase_",
  names(combined_df),
  value = TRUE
)
timing_cols <- timing_cols[!grepl("PGQ", timing_cols)]

# Apply the offset per column using the year embedded at the end of the column name
for (col in timing_cols) {
  yr <- as.integer(sub(".*_", "", col))
  if (is.na(yr)) next  # skip if the name doesn't end in _YYYY

  x <- combined_df[[col]]

  combined_df[[col]] <- ifelse(!is.na(x) & x > 366, x - 366 * (yr - 2000), x) # only correct values that exceed the DOY range
}
```

---

## Define a helper function to read and standardise a single climate CSV from GEE. The variable name is taken from the filename and used as the column name.`cell` is set as the key for merging

```{r}
read_one_clim <- function(f) {
  dt <- fread(f)
  dt[, cell := as.integer(cell)]
  var_col <- setdiff(names(dt), "cell")
  dt[, (var_col) := as.numeric(get(var_col))]
  dt <- dt[, c("cell", var_col), with = FALSE]
  setkey(dt, cell)
  dt
}
```

---

## Read and merge all 19 climate CSVs into a single wide-format table, joining on `cell`.

```{r}
climfiles <- list.files(clim_folder, pattern = "\\.csv$", full.names = TRUE)

for (f in climfiles) {
  cl <- read_one_clim(f)
  combined_df <- merge(combined_df, cl, by = "cell", all.x = TRUE)
}

dim(combined_df)
```

---

## Rescale temperature bioclimate variables from their raw integer encoding. WorldClim temperature variables are stored as integers scaled by a factor of 10 (or 100 for bio04); dividing restores values to degrees Celsius and appropriate units.

```{r}
temp_vars <- c("bio01","bio02","bio05","bio06","bio07","bio08","bio09","bio10","bio11")
combined_df[, (temp_vars) := lapply(.SD, function(x) x * 0.1), .SDcols = temp_vars]
combined_df[, bio04 := bio04 * 0.01]
```

---

## Rename all bioclimate and phenology columns to human-readable labels. A lookup table maps bioclimate variable codes to their standard names; phenology column names are cleaned by stripping the GEE prefix and converting underscores to title case.

```{r}
# Bio variable label lookup
bio_labels <- c(
  "01" = "Annual Mean Temp",
  "02" = "Mean Diurnal Range",
  "03" = "Isothermality",
  "04" = "Temp Seasonality",
  "05" = "Max Temp Warmest Month",
  "06" = "Min Temp Coldest Month",
  "07" = "Temp Annual Range",
  "08" = "Mean Temp Wettest Quarter",
  "09" = "Mean Temp Driest Quarter",
  "10" = "Mean Temp Warmest Quarter",
  "11" = "Mean Temp Coldest Quarter",
  "12" = "Annual Precipitation",
  "13" = "Precip Wettest Month",
  "14" = "Precip Driest Month",
  "15" = "Precip Seasonality",
  "16" = "Precip Wettest Quarter",
  "17" = "Precip Driest Quarter",
  "18" = "Precip Warmest Quarter",
  "19" = "Precip Coldest Quarter"
)

rename_var <- function(v) {
  # Bio variable rename
  num <- regmatches(v, regexpr("(?<=bio)\\d+", v, perl = TRUE))
  if (length(num) == 1 && num %in% names(bio_labels)) {
    return(paste0("BIO_", num, " ", bio_labels[[num]]))
  }

  # Phenology variable rename
  v <- gsub("gfbi_vnp22q2_", "", v)
  v <- gsub("_", " ", v)
  gsub("(^|\\s)(\\w)", "\\1\\U\\2", v, perl = TRUE)
}

# Identify all bio and phenology columns and apply the rename function
cols_to_rename <- grep("^bio\\d+$|^gfbi_vnp22q2_", names(combined_df), value = TRUE)

setnames(combined_df,
         old = cols_to_rename,
         new = sapply(cols_to_rename, rename_var))
```

---

## Export the combined dataframe as a CSV for use in downstream modelling.

```{r}
fwrite(combined_df, '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Raw_Combined_Data.csv')
```
