---
title: "Creating a Modelling Dataframe"
output: html_document
---

This notebook prepares the modelling-ready dataset from the summary dataframe produced by the *Creating a Summary Dataframe* notebook. Two cleaning steps are applied: the `n_stems_total` column is removed as it is not used as a predictor, and all rows containing any NA values are dropped to produce a complete-case dataset.

The number of rows removed and remaining are printed for reference. The final dataframe is exported as a CSV for use in model fitting.

---

## Load the packages required for data manipulation.

```{r}
library(data.table)
library(dplyr)
```

---

## Read in the summary dataframe produced by the *Creating a Summary Dataframe* notebook.

```{r}
model_df <- fread('/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Summary_Combined_Data.csv')
```

---

## Drop the `n_stems_total` column as it is not required for modelling.

```{r}
model_df[, n_stems_total := NULL]
```

---

## Remove all rows containing any NA values to produce a complete-case dataset. Row counts before and after are printed for reference.

```{r}
rows_before <- nrow(model_df)

model_df <- na.omit(model_df)

rows_after <- nrow(model_df)
cat("Rows removed:", rows_before - rows_after, "\n")
cat("Rows remaining:", rows_after, "\n")
```

---

## Export the modelling dataframe as a CSV for use in model fitting.

```{r}
fwrite(model_df, '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Modelling_Data.csv')
```
