---
title: "Combined Model - Variable Importance Plot"
output: html_document
---

This notebook reads the variable importance summary produced by the *Combined Model - Variable Importance Summary* notebook and produces an importance plot for the top variables. The raw per-run importance scores are also read in to provide the jittered run-level data points.

Variable labels on the y-axis are colour-coded by type: blue for climate (BIO) variables and green for phenology variables. Individual run importance values are shown as jittered points, with mean ± SD overlaid in red. The plot is exported to the importance ranks output directory.

---

## Load the packages required for data handling and plotting.

```{r}
library(data.table)
library(ggplot2)
library(ggtext)
```

---

## Read in the variable importance summary and the raw per-run importance scores. The summary provides the mean and SD used for the error bars; the raw scores are used for the jittered points.

```{r}
pull_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
save_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/ImportanceRanks'
imp_summary <- fread(file.path(save_dir, "CombinedModelImportanceSummary.csv"))
results_imp <- fread(file.path(pull_dir, "AllCombinedModelImportance.csv"))
```

---

## Rename phenology variables to remove the season indicator ' 1 ' and drop ' Mean' from mean variable names, leaving SD variables labelled with ' SD' only.

```{r}
rename_pheno <- function(v) {
  is_pheno <- !grepl("^BIO_", v)
  v[is_pheno] <- gsub(" 1 Mean$", "",    v[is_pheno])  # "X 1 Mean" → "X"
  v[is_pheno] <- gsub(" 1 SD$",   " SD", v[is_pheno])  # "X 1 SD"   → "X SD"
  v
}

imp_summary[, variable := rename_pheno(variable)]
results_imp[, variable := rename_pheno(variable)]
```

---

## Define a helper function to clean variable names for axis labels. Removes the BIO_ prefix from climate variables.

```{r}
clean_var_name <- function(v) gsub("^BIO_\\d+\\s+", "", v)
```

---

## Build the variable importance plot for the top variables. Jittered points show individual run values in original permutation importance units; red points and error bars show mean ± 95% CI across runs. Y-axis labels are coloured by variable type and rendered via `ggtext`.

```{r fig.width=14, fig.height=12}
# Subset to top variables and extract their per-run values for jittering
most_imp        <- head(imp_summary, 57)
most_imp_jitter <- results_imp[variable %in% most_imp$variable]

# Label each variable as Climate or Phenology for colour mapping in the legend
most_imp_jitter[, var_type := ifelse(grepl("^BIO_", variable), "Climate", "Phenology")]

# Build coloured axis labels: BIO variables in blue, phenology in green
# Rank is shown in brackets at the start; phenology variables are prefixed with *
# Labels are wrapped in a <span> tag so ggtext can render the colour in the axis
make_html_label <- function(v, rank) {
  is_pheno <- !grepl("^BIO_", v)
  colour   <- ifelse(is_pheno, "#4caf50", "blue")
  prefix   <- ""
  paste0("<span style='color:black; font-weight:bold'>(", rank, ") </span>",
         "<span style='color:", colour, "; font-weight:bold'>", prefix, clean_var_name(v), "</span>")
}

# Build a named lookup so both most_imp and most_imp_jitter get identical labels
label_lookup <- setNames(
  mapply(make_html_label, most_imp$variable, most_imp$rank),
  most_imp$variable
)

# Order variables ascending by importance so that coord_flip() places the most
# important variable at the top; set as a factor so all geoms share the same axis order
var_order   <- most_imp[order(mean_importance, decreasing = TRUE), variable]
html_levels <- label_lookup[var_order]

most_imp[,        label_html := factor(label_lookup[variable], levels = html_levels)]
most_imp_jitter[, label_html := factor(label_lookup[variable], levels = html_levels)]


importance_plot <- ggplot() +
  # Jittered points show per-run permutation importance in original units
  geom_jitter(data = most_imp_jitter,
              aes(x = label_html, y = perm_importance, colour = var_type),
              width = 0.2, alpha = 0.3, size = 1) +
  # Error bars span mean ± 95% CI in original permutation importance units
  geom_errorbar(data = most_imp,
                aes(x = label_html,
                    ymin = mean_importance - ci95_importance,
                    ymax = mean_importance + ci95_importance),
                colour = "black", width = 0.5, linewidth = 1.2) +
  # Red point marks the mean permutation importance for each variable
  geom_point(data = most_imp,
             aes(x = label_html, y = mean_importance),
             colour = "red", size = 1) +
  # Colour scale maps variable type to the same colours used in the axis labels
  scale_colour_manual(values = c("Climate" = "steelblue", "Phenology" = "#4caf50"),
                      name = "Variable type") +
  # Override legend aesthetics so points are visible at full opacity
  guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  scale_y_continuous(limits = c(0, NA)) +
  coord_flip() +
  labs(x = NULL, y = "Mean Permutation Importance ± 95% CI") +
  theme_minimal() +
  # axis.text.y uses element_markdown to render the colour tags in axis labels
  theme(
    plot.title        = element_text(size = 15),
    axis.title        = element_text(size = 15),
    axis.text.x       = element_text(size = 13),
    axis.text.y       = element_markdown(size = 14),
    legend.title      = element_text(size = 15),
    legend.text       = element_text(size = 14),
    axis.line         = element_line(colour = "black", linewidth = 0.5),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(3, "pt")
  )

print(importance_plot)
```

---

## Export the variable importance plot as a PNG.

```{r}
ggsave(file.path(save_dir, "CombinedModel_ImportancePlot.png"),
       plot   = importance_plot,
       width  = 14,
       height = 12)
```
