---
title: "Combined Model - Variable Importance Summary"
output: html_document
---

This notebook loads the per-run permutation importance scores from the 100-run combined (climate + phenology) random forest produced by the *Climate & Phenology Combined Ranger Models - 100 Run Loop* notebook, and summarises mean and SD of permutation importance per variable across all 100 runs. The summary table is exported as a CSV for use in the *Combined Model - Variable Importance Plot* notebook.

---

## Load the packages required for data handling.

```{r}
library(data.table)
```

---

## Read in the saved per-run variable importance scores from the 100-run combined model.

```{r}
pull_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
save_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/ImportanceRanks'
results_imp <- fread(file.path(pull_dir, "AllCombinedModelImportance.csv"))
```

---

## Summarise variable importance across all 100 runs by computing mean, SD, and 95% CI of permutation importance per variable, ordered from most to least important. A rank column is assigned based on this order.

```{r}
imp_summary <- results_imp[, .(
  mean_importance  = mean(perm_importance),
  sd_importance    = sd(perm_importance),
  ci95_importance  = qt(0.975, df = .N - 1) * sd(perm_importance) / sqrt(.N)
), by = variable][order(mean_importance, decreasing = TRUE)]

imp_summary[, rank := .I]

print(imp_summary)
```

---

## Export the variable importance summary table as a CSV.

```{r}
fwrite(imp_summary, file.path(save_dir, "CombinedModelImportanceSummary.csv"))
```
