---
title: "Combined Model — Backward Elimination Variable Importance"
output: html_document
---

This notebook performs backward elimination on the combined (climate + phenology) random forest predictor set. Starting from the full 57-variable model, variables are removed one batch at a time in reverse importance order (least important first), using the importance ranks from the *Combined Model — Variable Importance Summary* notebook. A single ranger model is fit at each elimination step using a randomly selected seed from the climate model loop. OOB R² and RMSE are recorded at every step to identify the variable subset that maximises predictive performance.

The removal schedule checks at: 57 → 47 → 37 → 32 → 30, then removes one variable at a time from 30 down to 1.

---

## Load the packages required for modelling, data handling, and plotting.

```{r}
library(data.table)
library(ranger)
library(ggplot2)
```

---

## Define file paths for input data and output artefacts.

```{r}
model_dir  <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
data_dir   <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing'
imp_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/ImportanceRanks'
```

---

## Load the modelling dataset, climate model run statistics, and combined model importance summary.

```{r}
model_df    <- fread(file.path(data_dir, "Modelling_Data.csv"))
clim_stats  <- fread(file.path(model_dir, "AllClimateModelStats.csv"))
imp_summary <- fread(file.path(imp_dir,   "CombinedModelImportanceSummary.csv"))
```

---

## Select a seed at random from the climate model loop runs.

```{r}
random_run   <- clim_stats[sample(.N, 1)]
elim_seed <- random_run$seed

cat(sprintf("Selected run: %d  |  seed: %d  |  R² = %.4f\n",
            rep_run$run, elim_seed, rep_run$r_squared))
```

---

## Build the ordered predictor list: all combined-model variables sorted from least to most important. The backward elimination loop will remove from the front of this list.

```{r}
# Full combined predictor set (same definition as CombinedRangerModelLoop.Rmd)
all_vars <- setdiff(names(model_df), c("cell", "propEcM"))

# Order importance summary to match and sort worst-first (highest rank number = least important)
imp_ordered <- imp_summary[variable %in% all_vars][order(-rank)]

# Verify all predictors are accounted for in the importance summary
if (nrow(imp_ordered) != length(all_vars)) {
  warning(sprintf(
    "%d predictors in model_df but %d in importance summary — check for mismatches.",
    length(all_vars), nrow(imp_ordered)
  ))
}

# Variables in elimination order: first element is removed first (least important)
elim_order <- imp_ordered$variable

cat(sprintf("Total predictors: %d\n", length(elim_order)))
cat("First 5 to be removed (least important):\n")
print(head(elim_order, 5))
cat("Last 5 to be removed (most important):\n")
print(tail(elim_order, 5))
```

---

## Define the elimination checkpoints as the number of variables remaining at each step.

```{r}
n_vars <- length(elim_order)

# Variables remaining at each checkpoint: 57→47→37→32→30, then 29 down to 1 one at a time
remaining_schedule <- c(57, 47, 37, 32, 30, seq(29, 1, by = -1))

# Convert to number removed at each checkpoint
checkpoints <- as.integer(n_vars - remaining_schedule)

cat(sprintf("Elimination checkpoints (n remaining): %s\n",
            paste(remaining_schedule, collapse = ", ")))
cat(sprintf("Total model fits: %d\n", length(checkpoints)))
```

---

## Run a single ranger model at each elimination checkpoint. At each step the least-important remaining variables are dropped and the model is re-fit on the reduced predictor set.

```{r}
ntree  <- 500
ncores <- 6

elim_results <- vector("list", length(checkpoints))

for (k in seq_along(checkpoints)) {

  n_removed     <- checkpoints[k]
  active_vars   <- if (n_removed == 0) elim_order else elim_order[-(1:n_removed)]
  n_active      <- length(active_vars)

  m <- ranger(
    x            = model_df[, ..active_vars],
    y            = model_df$propEcM,
    num.trees    = ntree,
    num.threads  = ncores,
    write.forest = FALSE,
    seed         = elim_seed
  )

  elim_results[[k]] <- data.frame(
    step       = k,
    n_removed  = n_removed,
    n_vars     = n_active,
    r_squared  = m$r.squared,
    mse        = m$prediction.error,
    rmse       = sqrt(m$prediction.error)
  )

  message(sprintf("Step %2d/%d  |  %d vars remaining  |  R² = %.4f  |  RMSE = %.5f",
                  k, length(checkpoints), n_active,
                  m$r.squared, sqrt(m$prediction.error)))
}

elim_dt <- rbindlist(elim_results)
```

---

## Report the optimal variable subsets: the step with the highest OOB R² and the step with the lowest OOB RMSE.

```{r}
best_r2   <- elim_dt[which.max(r_squared)]
best_rmse <- elim_dt[which.min(rmse)]

cat("Optimal subset by highest R²:\n")
cat(sprintf("  Variables remaining: %d  |  R² = %.4f  |  RMSE = %.5f\n\n",
            best_r2$n_vars, best_r2$r_squared, best_r2$rmse))

cat("Optimal subset by lowest RMSE:\n")
cat(sprintf("  Variables remaining: %d  |  R² = %.4f  |  RMSE = %.5f\n\n",
            best_rmse$n_vars, best_rmse$r_squared, best_rmse$rmse))

# Variables retained at the best-R² step
n_kept_r2   <- best_r2$n_removed
active_at_best_r2 <- if (n_kept_r2 == 0) elim_order else elim_order[-(1:n_kept_r2)]
cat(sprintf("Variables retained at best-R² step (%d vars):\n", length(active_at_best_r2)))
print(active_at_best_r2)
```

---

## Save the per-step results table as a CSV.

```{r}
fwrite(elim_dt, file.path(imp_dir, "BackwardElimination_Results.csv"))
```

---

## Define the shared plot function and phase boundaries.

```{r}
# Phase boundary x-positions (n_vars remaining at each step-size change)
phase_boundaries <- elim_dt$n_vars[elim_dt$n_removed %in% c(10L, 20L, 25L, 27L)]

make_elim_plot <- function(data, y_col, y_label, best_row, fill_col, point_col) {

  best_x <- best_row$n_vars
  best_y <- best_row[[y_col]]

  annot_label <- sprintf("n = %d\n%s = %.4f", best_x, y_label, best_y)

  nudge_y <- if (y_col == "r_squared") -diff(range(data[[y_col]])) * 0.10
             else                        diff(range(data[[y_col]])) * 0.10

  ggplot(data, aes(x = n_vars, y = .data[[y_col]])) +

    geom_vline(xintercept = phase_boundaries,
               linetype = "dashed", colour = "grey60", linewidth = 0.5) +

    geom_line(colour = "grey30", linewidth = 0.7) +
    geom_point(colour = "grey30", fill = fill_col,
               size = 2.5, shape = 21, stroke = 0.5) +

    geom_point(data = best_row, aes(x = n_vars, y = .data[[y_col]]),
               colour = point_col, fill = point_col,
               size = 4, shape = 21, stroke = 1.2) +
    annotate("text",
             x = best_x, y = best_y + nudge_y,
             label = annot_label,
             hjust = 0.5, size = 3.5, colour = point_col, lineheight = 1.3) +

    scale_x_continuous(
      name   = "Number of variables",
      breaks = data$n_vars
    ) +
    labs(y = y_label) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      axis.title       = element_text(size = 11),
      axis.text.x      = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8)
    )
}
```

---

## Plot OOB R² as a function of the number of variables remaining.

```{r fig.width=12, fig.height=5}
p_r2 <- make_elim_plot(
  data      = elim_dt,
  y_col     = "r_squared",
  y_label   = "OOB R²",
  best_row  = best_r2,
  fill_col  = "#cce5ff",
  point_col = "steelblue4"
)

p_r2
```

---

## Plot OOB RMSE as a function of the number of variables remaining.

```{r fig.width=12, fig.height=5}
p_rmse <- make_elim_plot(
  data      = elim_dt,
  y_col     = "rmse",
  y_label   = "OOB RMSE",
  best_row  = best_rmse,
  fill_col  = "#ffd6cc",
  point_col = "firebrick3"
)

p_rmse
```

---

## Save the R² and RMSE plots as PNGs.

```{r}
ggsave(
  filename = file.path(imp_dir, "BackwardElimination_R2.png"),
  plot     = p_r2,
  width    = 12,
  height   = 5,
  dpi      = 300
)

ggsave(
  filename = file.path(imp_dir, "BackwardElimination_RMSE.png"),
  plot     = p_rmse,
  width    = 12,
  height   = 5,
  dpi      = 300
)
```
