---
title: "GFBI Cell Location Map"
output: html_document
---

This notebook maps the geographic locations of all GFBI grid cells used in the analysis, plotted over a North America basemap. Cell coordinates are sourced from the GFBI VIIRS export in WGS84 lon/lat. The dataset is restricted to cells present in the simple mycorrhizal dataset by joining on cell ID.

The resulting map provides a visual reference for the geographic coverage and spatial distribution of the GFBI data across North America. The plot is exported as a PNG.

---

## Load the packages required for spatial data handling and plotting.

```{r}
library(data.table)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
```

---

## Define file paths for input data and plot output directories.

```{r}
data_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing'
gee_dir  <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Google Earth Engine'
save_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/SupplementaryPlots'
```

---

## Load cell coordinates and restrict to cells present in the simple mycorrhizal dataset by joining on cell ID.

```{r}
coords     <- fread(file.path(gee_dir, "GFBI_VIIRS_Cells_Data_1.csv"))
simple_myc <- fread(file.path(data_dir, "Simple_Myc_Data.csv"))

# Retain only cells present in the simple mycorrhizal dataset
cells <- coords[simple_myc[, .(cell)], on = "cell", nomatch = 0]
```

---

## Build the North America basemap using Natural Earth country boundaries.

```{r}
north_america <- ne_countries(
  continent   = "North America",
  scale       = "medium",
  returnclass = "sf"
)
```

---

## Plot cell locations over the basemap. Each point represents one GFBI grid cell.

```{r fig.width=11, fig.height=7}
p_locations <- ggplot() +
  geom_sf(data = north_america, fill = "grey90", colour = "grey60", linewidth = 0.3) +
  geom_point(
    data   = cells,
    aes(x = lon, y = lat),
    colour = "steelblue", size = 0.1, alpha = 0.8
  ) +
  coord_sf(
    xlim   = c(-180, -40),
    ylim   = c(7, 80),
    expand = FALSE,
    datum  = sf::st_crs(4326)
  ) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  labs(
    x     = "Longitude",
    y     = "Latitude"
  ) +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(
    location = "br", which_north = "true",
    style    = north_arrow_fancy_orienteering()
  ) +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_locations
```

---

## Export the cell location map as a PNG.

```{r}
ggsave(
  file.path(save_dir, "gfbi_cell_locations.png"),
  plot   = p_locations,
  width  = 11,
  height = 7
)
```

```{r}
citation()
citation("ggplot2")
citation("sf")
citation("rnaturalearth")
citation("rnaturalearthdata")
citation("ggspatial")
```

