---
title: "GFBI Cell EcM Proportion Map"
output: html_document
---

This notebook maps the geographic locations of all GFBI grid cells used in the analysis, coloured by the proportion of ectomycorrhizal (EcM) associated trees (`propEcM`). Cell coordinates are sourced from the GFBI VIIRS export in WGS84 lon/lat. `propEcM` is joined from the simple mycorrhizal dataset by cell ID.

The plasma colour scale runs from high EcM proportion (light yellow) to low (dark purple), reversed so that high values are visually prominent. The map provides a spatial reference for the geographic distribution of EcM dominance across the North American GFBI data. The plot is exported as a PNG.

---

## Load the packages required for spatial data handling and plotting.

```{r}
library(data.table)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
```

---

## Define file paths for input data and plot output directories.

```{r}
data_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing'
gee_dir  <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Google Earth Engine'
save_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/SupplementaryPlots'
```

---

## Load cell coordinates and join propEcM from the simple mycorrhizal dataset by cell ID.

```{r}
coords     <- fread(file.path(gee_dir, "GFBI_VIIRS_Cells_Data_1.csv"))
simple_myc <- fread(file.path(data_dir, "Simple_Myc_Data.csv"))

# Join propEcM onto coordinates; retain only cells present in the simple mycorrhizal dataset
cells <- coords[simple_myc[, .(cell, propEcM)], on = "cell", nomatch = 0]
```

---

## Build the North America basemap using Natural Earth country boundaries.

```{r}
north_america <- ne_countries(
  continent   = "North America",
  scale       = "medium",
  returnclass = "sf"
)
```

---

## Plot cell locations coloured by propEcM over the basemap. The plasma scale is reversed so that high EcM proportion appears in light yellow.

```{r fig.width=10, fig.height=8}
p_ecm <- ggplot() +
  geom_sf(data = north_america, fill = "grey90", colour = "grey60", linewidth = 0.3) +
  geom_point(
    data = cells,
    aes(x = lon, y = lat, colour = propEcM),
    size = 0.1, alpha = 0.8
  ) +
  coord_sf(
    xlim   = c(-180, -40),
    ylim   = c(7, 80),
    expand = FALSE,
    datum  = sf::st_crs(4326)
  ) +
  scale_y_continuous(breaks = seq(10, 80, by = 10)) +
  scale_colour_viridis_c(
    name      = "propEcM",
    limits    = c(0, 1),
    option    = "plasma",
    direction = -1
  ) +
  labs(
    x     = "Longitude",
    y     = "Latitude"
  ) +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(
    location = "br", which_north = "true",
    style    = north_arrow_fancy_orienteering()
  ) +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_ecm
```

---

## Export the EcM proportion map as a PNG.

```{r}
ggsave(
  file.path(save_dir, "gfbi_cell_ecm_proportion.png"),
  plot   = p_ecm,
  width  = 10,
  height = 8
)
```
