---
title: "Phenology Model - Hyperparameter Tuning"
output: html_document
---

This notebook tunes three key hyperparameters of the phenology-only random forest model: `num.trees`, `mtry`, and `min.node.size`. The phenology model is the focal model for this study and is therefore the priority for hyperparameter evaluation. Each parameter is tuned sequentially: `num.trees` first to establish a stable ensemble size, then `mtry` (the most influential parameter) using the chosen tree count, and finally `min.node.size`. A single fixed seed is used throughout to isolate the effect of each parameter. OOB R² is used as the performance metric throughout, as it provides an honest estimate of generalisation without requiring a separate validation set.

---

## Load required packages.

```{r}
library(data.table)
library(ranger)
library(ggplot2)
```

---

## Read in the modelling dataset and define the phenology-only predictor set.

```{r}
model_df <- fread('/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Modelling_Data.csv')

set.seed(42)
model_df <- model_df[sample(.N, 50000)]

predictor_vars <- setdiff(names(model_df), c("cell", "propEcM"))
predictor_vars <- predictor_vars[!grepl("^BIO_", predictor_vars)]

cat("Number of phenology predictors:", length(predictor_vars), "\n")
cat("Default mtry (floor(p/3)):", floor(length(predictor_vars) / 3), "\n")
```

---

## Step 1 — Tune `num.trees`.

OOB R² is recorded across a range of tree counts. The plateau indicates the minimum ensemble size needed for stable predictions. A fixed seed is used so that any change in OOB R² reflects tree count alone.

```{r}
tree_vals <- c(50, 100, 150, 200, 300, 400, 500, 750, 1000)

tree_results <- rbindlist(lapply(tree_vals, function(nt) {
  m <- ranger(
    x             = model_df[, ..predictor_vars],
    y             = model_df$propEcM,
    num.trees     = nt,
    num.threads   = 7,
    seed          = 42
  )
  data.frame(num_trees = nt, r_squared = m$r.squared)
}))

print(tree_results)
```

```{r}
ggplot(tree_results, aes(x = num_trees, y = r_squared)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.8) +
  geom_point(colour = "#2c7bb6", size = 2.5) +
  geom_vline(xintercept = 500, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 510, y = min(tree_results$r_squared),
           label = "Default", hjust = 0, colour = "grey50", size = 3.5) +
  labs(
    x     = "Number of Trees",
    y     = expression("OOB R"^2),
    title = "Phenology Model: OOB R² vs Number of Trees"
  ) +
  theme_classic()

ggsave(
  filename = file.path(
    '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/TuningParams',
    "PhenoModel_Trees_Tuning.png"
  ),
  width = 7, height = 5, dpi = 300
)
```

---

## Step 2 — Tune `mtry`.

`mtry` controls the number of predictors randomly considered at each split and is the most influential hyperparameter in random forest models. With 38 phenology predictors, the default is `floor(38/3) = 12`. A range spanning well below and above this default is tested. `num.trees` is fixed at 500 throughout.

```{r}
mtry_vals <- c(2, 4, 6, 8, 10, 12, 16, 20, 25, 30, 38)

mtry_results <- rbindlist(lapply(mtry_vals, function(mt) {
  m <- ranger(
    x             = model_df[, ..predictor_vars],
    y             = model_df$propEcM,
    num.trees     = 500,
    mtry          = mt,
    num.threads   = 7,
    seed          = 42
  )
  data.frame(mtry = mt, r_squared = m$r.squared)
}))

print(mtry_results)
```

```{r}
ggplot(mtry_results, aes(x = mtry, y = r_squared)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.8) +
  geom_point(colour = "#2c7bb6", size = 2.5) +
  geom_vline(xintercept = floor(length(predictor_vars) / 3),
             linetype = "dashed", colour = "grey50") +
  annotate("text", x = floor(length(predictor_vars) / 3) + 0.5,
           y = min(mtry_results$r_squared),
           label = "Default", hjust = 0, colour = "grey50", size = 3.5) +
  labs(
    x     = "mtry",
    y     = expression("OOB R"^2),
    title = "Phenology Model: OOB R² vs mtry"
  ) +
  theme_classic()

ggsave(
  filename = file.path(
    '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/TuningParams',
    "PhenoModel_Mtry_Tuning.png"
  ),
  width = 7, height = 5, dpi = 300
)
```

---

## Step 3 — Tune `min.node.size`.

`min.node.size` sets the minimum number of observations required in a terminal node, controlling tree depth. Smaller values allow deeper, more flexible trees; larger values produce shallower, more regularised trees. The default for regression is 5. `num.trees` is fixed at 500 and `mtry` at the ranger default throughout.

```{r}
node_vals <- c(1, 2, 5, 10, 20, 30, 50, 75, 100)

node_results <- rbindlist(lapply(node_vals, function(ns) {
  m <- ranger(
    x             = model_df[, ..predictor_vars],
    y             = model_df$propEcM,
    num.trees     = 500,
    min.node.size = ns,
    num.threads   = 7,
    seed          = 42
  )
  data.frame(min_node_size = ns, r_squared = m$r.squared)
}))

print(node_results)
```

```{r}
ggplot(node_results, aes(x = min_node_size, y = r_squared)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.8) +
  geom_point(colour = "#2c7bb6", size = 2.5) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 5.5, y = min(node_results$r_squared),
           label = "Default", hjust = 0, colour = "grey50", size = 3.5) +
  labs(
    x     = "Minimum Node Size",
    y     = expression("OOB R"^2),
    title = "Phenology Model: OOB R² vs Minimum Node Size"
  ) +
  theme_classic()

ggsave(
  filename = file.path(
    '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/TuningParams',
    "PhenoModel_NodeSize_Tuning.png"
  ),
  width = 7, height = 5, dpi = 300
)
```

---

