---
title: "Climate Model - Predicted vs Observed Plot"
output: html_document
---

This notebook loads the OOB predictions and model statistics from the 100-run climate random forest produced by the *Climate Ranger Models - 100 Run Loop* notebook, and produces a predicted vs observed plot for `propEcM`. Per-cell mean and SD of predicted `propEcM` are computed across all 100 runs before plotting.

The plot uses a hexbin density layer to handle the large number of cells, with a fitted regression line and a 1:1 reference line overlaid. Mean R² and RMSE across the 100 runs are annotated on the plot. A binned RMSE plot is also produced, with separate bins for cells with observed propEcM of exactly 0 or 1, and 0.05-width bins for intermediate values. The figures are exported as PNGs to the model outputs directory.

---

## Load the packages required for data handling and plotting.

```{r}
library(data.table)
library(ggplot2)
library(hexbin)
library(geomtextpath)
```

---

## Read in the saved model statistics and OOB predictions from the 100-run climate model.

```{r}
save_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'

results_stats <- fread(file.path(save_dir, "AllClimateModelStats.csv"))
results_preds <- fread(file.path(save_dir, "AllClimateModelPredictions.csv"))
```

---

## Compute per-cell mean and SD of predicted `propEcM` across all 100 runs, and derive summary model performance statistics.

```{r}
# Aggregate predictions to per-cell means and SDs across runs
mean_preds <- results_preds[, .(
  observed     = mean(observed),
  predicted    = mean(predicted),
  predicted_sd = sd(predicted)
), by = cell]

# Compute mean R² and RMSE across the 100 runs for annotation
r2   <- round(mean(results_stats$r_squared), 3)
rmse <- round(sqrt(mean(results_stats$prediction_error)), 3)

# Compute agreement statistics on per-cell mean predictions
lm_fit    <- lm(predicted ~ observed, data = mean_preds)
slope     <- round(coef(lm_fit)[2], 3)
intercept <- round(coef(lm_fit)[1], 3)
bias      <- round(mean(mean_preds$predicted - mean_preds$observed), 3)

eq_label <- paste0("lm  y = ", slope, "x ", ifelse(intercept >= 0, "+ ", "- "), abs(intercept))
```

---

## Bin observed values with separate bins for exactly 0 and 1, and 0.05-width bins for intermediates, then compute RMSE per bin and plot.

```{r}
breaks_interior <- seq(0, 1, by = 0.05)

mean_preds[, obs_bin := fcase(
  observed == 0, "0",
  observed == 1, "1",
  default = as.character(cut(observed, breaks = breaks_interior,
                             include.lowest = FALSE, right = TRUE))
)]

# Assign the same bins to raw per-run predictions
results_preds[, obs_bin := fcase(
  observed == 0, "0",
  observed == 1, "1",
  default = as.character(cut(observed, breaks = breaks_interior,
                             include.lowest = FALSE, right = TRUE))
)]

# RMSE per bin per run, then summarise across 100 runs
bin_rmse_by_run <- results_preds[, .(
  RMSE = sqrt(mean((predicted - observed)^2))
), by = .(obs_bin, run)]

bin_meta <- mean_preds[, .(bin_mid = mean(observed), n_cells = .N), by = obs_bin]

bin_errors <- bin_rmse_by_run[, .(
  RMSE    = mean(RMSE),
  RMSE_sd = sd(RMSE),
  CI95_lo = mean(RMSE) - qt(0.975, df = .N - 1) * sd(RMSE) / sqrt(.N),
  CI95_hi = mean(RMSE) + qt(0.975, df = .N - 1) * sd(RMSE) / sqrt(.N)
), by = obs_bin]

bin_errors <- merge(bin_errors, bin_meta, by = "obs_bin")[order(bin_mid)]
bin_errors[, is_boundary := obs_bin %in% c("0", "1")]

binned_error_plot <- ggplot(bin_errors, aes(x = bin_mid, y = RMSE)) +
  geom_col(data = bin_errors[is_boundary == FALSE],
           width = 0.038, fill = "#7b2d8b", colour = NA) +
  geom_line(linewidth = 0.8, colour = "#f28c50") +
  geom_point(size = 3, shape = 18, colour = "#f28c50") +
  geom_errorbar(aes(ymin = CI95_lo, ymax = CI95_hi),
                width = 0.015, linewidth = 0.4, colour = "#f28c50") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  coord_cartesian(ylim = c(0, 0.4)) +
  labs(x = "Observed propEcM",
       y = "RMSE") +
  theme_minimal() +
  theme(
    axis.line         = element_line(colour = "black", linewidth = 0.4),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(3, "pt")
  )

print(binned_error_plot)

ggsave(file.path(save_dir, "ClimateModel_BinnedRMSEPlot.png"),
       plot   = binned_error_plot,
       width  = 9,
       height = 6)
```

---

## Compute a per-bin RMSE table with SD and 95% CI across the 100 runs, and export as CSV.

```{r}
knitr::kable(
  bin_errors[, .(
    Bin     = obs_bin,
    n_cells = n_cells,
    RMSE    = round(RMSE,    4),
    SD      = round(RMSE_sd, 4),
    CI95_lo = round(CI95_lo, 4),
    CI95_hi = round(CI95_hi, 4)
  )],
  caption = "Per-bin RMSE — mean, SD, and 95% CI across 100 climate model runs"
)

fwrite(
  bin_errors[, .(obs_bin, bin_mid, n_cells, RMSE, RMSE_sd, CI95_lo, CI95_hi)],
  file.path(save_dir, "ClimateModel_BinnedRMSETable.csv")
)
```

---

## Build the predicted vs observed plot. Hexbin density is used to handle the large number of cells. The dashed red line is the 1:1 ideal and the white line is the fitted linear model. R² and RMSE are annotated in the plot area.

```{r}
pred_obs_plot <- ggplot(mean_preds, aes(x = observed, y = predicted)) +
  geom_hex(bins = 80) +
  scale_fill_viridis_c(option = "magma", name = "No. of Cells ", trans = "log10") +
  geom_textsmooth(method = "lm", label = eq_label, colour = "white",
                linewidth = 0.8, se = FALSE, hjust = 0.05, vjust = -0.5) +
  geom_textabline(intercept = 0, slope = 1, label = "1:1",
                colour = "red", linetype = "dashed", linewidth = 0.8, hjust = 0.123, vjust = -0.5) +
  annotate("text", x = 0.68, y = 0.38,
           label = paste0("R² = ", r2, "\nRMSE = ", rmse),
           hjust = 0, vjust = 0, size = 4, colour = "white") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), minor_breaks = seq(0, 1, 0.05)) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Observed propEcM",
       y = "Mean Predicted propEcM (100 runs)") +
  theme_minimal() +
  theme(
    axis.line         = element_line(colour = "black", linewidth = 0.4),
    axis.ticks        = element_line(colour = "black"),
    axis.ticks.length = unit(3, "pt")
  )

print(pred_obs_plot)
```

---

## Export the predicted vs observed plot as a PNG to the model outputs directory.

```{r}
ggsave(file.path(save_dir, "ClimateModel_PredVsObsPlot.png"),
       plot   = pred_obs_plot,
       width  = 9,
       height = 6)
```


```{r}
citation("hexbin")
citation("geomtextpath")
```

