---
title: "Phenology Ranger Models - 100 Run Loop"
output: html_document
---

This notebook fits 100 independent random forest models using the `ranger` package, each predicting ectomycorrhizal proportion (`propEcM`) from the 38 phenology summary variables (mean and SD per metric across 2013–2022). Climate variables are deliberately excluded so that the predictive power of phenology alone can be assessed. The input is the modelling dataframe produced by the *Creating a Modelling Dataframe* notebook.

Each model is fit with a unique reproducible seed drawn from a fixed random sequence. Running multiple independent models with different seeds captures variability in out-of-bag (OOB) performance arising from the stochastic nature of random forest fitting, providing a more stable estimate of model behaviour than a single run. For each run, per-model summary statistics and OOB predictions are collected and exported as CSVs.

---

## Load the packages required for modelling and data handling.

```{r}
library(data.table)
library(ranger)
```

---

## Read in the modelling dataset produced by the *Creating a Modelling Dataframe* notebook. All NA rows have been removed and columns are labelled — no further preprocessing is required.

```{r}
model_df <- fread('/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Modelling_Data.csv')
```

---

## Define the phenology-only predictor set by excluding the cell identifier, the response variable, and all BIO climate columns.

```{r}
predictor_vars <- setdiff(names(model_df), c("cell", "propEcM"))
predictor_vars <- predictor_vars[!grepl("^BIO_", predictor_vars)]
print(predictor_vars)
```

---

## Run 100 independent ranger models using the phenology predictor set. Each run uses a unique reproducible seed. Per-run summary statistics and OOB predictions are collected into lists and combined into dataframes at the end of the loop.

```{r}
# Set fixed model parameters
ntree      <- 500
ncores     <- 7
n_runs     <- 100
set.seed(22)
base_seeds <- sample.int(1e6, n_runs) # draw n_runs unique seeds for reproducibility

# Initialise storage lists for stats and predictions
run_stats  <- vector("list", n_runs)
pred_list  <- vector("list", n_runs)

for (i in seq_len(n_runs)) {

  seed_i <- base_seeds[i]

  m <- ranger(
    x            = model_df[, ..predictor_vars],
    y            = model_df$propEcM,
    num.trees    = ntree,
    num.threads  = ncores,
    write.forest = FALSE, # forest object not needed; suppressing saves memory
    seed         = seed_i
  )

  # Collect model-level summary statistics for this run
  run_stats[[i]] <- data.frame(
    run              = i,
    seed             = seed_i,
    r_squared        = m$r.squared,
    prediction_error = m$prediction.error,
    num_trees        = m$num.trees,
    mtry             = m$mtry
  )

  # Collect OOB predictions alongside observed values for this run
  pred_list[[i]] <- data.frame(
    run       = i,
    seed      = seed_i,
    cell      = model_df$cell,
    observed  = model_df$propEcM,
    predicted = m$predictions
  )

  if (i %% 10 == 0) message("Completed run ", i, "/", n_runs) # progress tracker
}

# Combine per-run results into single dataframes
results_stats <- rbindlist(run_stats)
results_preds <- rbindlist(pred_list)
```

---

## Export the per-run model statistics and OOB predictions as CSVs.

```{r}
save_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'

fwrite(results_stats, file.path(save_dir, "AllPhenoModelStats.csv"))
fwrite(results_preds, file.path(save_dir, "AllPhenoModelPredictions.csv"))
```

```{r}
citation("ranger")
```

