---
title: "Residual Ranger Models - 100 Run Loop"
output: html_document
---

This notebook fits 100 independent random forest models using the `ranger` package, each predicting climate-residual ectomycorrhizal proportion (`residualEcM`) from the 38 phenology summary variables only. The residual approach isolates the variance in `propEcM` that is not explained by climate: per-cell residuals are computed as observed `propEcM` minus mean predicted `propEcM` from the 100-run climate model produced by the *Climate Ranger Models - 100 Run Loop* notebook. These residuals become the response variable, with phenology as the sole predictor set.

Each model is fit with a unique reproducible seed drawn from a fixed random sequence. Running multiple independent models with different seeds captures variability in out-of-bag (OOB) performance arising from the stochastic nature of random forest fitting, providing a more stable estimate of model behaviour than a single run. Per-model summary statistics and OOB predictions are collected and exported as CSVs.

---

## Load the packages required for modelling and data handling.

```{r}
library(data.table)
library(ranger)
```

---

## Read in the modelling dataset and the saved OOB predictions from the 100-run climate model.

```{r}
model_df <- fread('/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Modelling_Data.csv')

save_dir    <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'

climate_preds <- fread(file.path(save_dir, "AllClimateModelPredictions.csv"))
```

---

## Compute per-cell mean predicted `propEcM` across all 100 climate model runs, then calculate `residualEcM` as observed minus mean predicted.

```{r}
# Average climate model predictions across runs for each cell
cell_means <- climate_preds[, .(
  mean_climate_predicted = mean(predicted),
  observed               = mean(observed)
), by = cell]

# Compute the residual as the difference between observed and climate-predicted propEcM
cell_means[, residualEcM := observed - mean_climate_predicted]
```

---

## Merge `residualEcM` into the modelling dataset so phenology predictors are aligned by cell.

```{r}
model_df <- merge(model_df, cell_means[, .(cell, residualEcM)], by = "cell")
```

---

## Define the phenology-only predictor set by excluding the cell identifier, both response variables, and all BIO climate columns.

```{r}
predictor_vars <- setdiff(names(model_df), c("cell", "propEcM", "residualEcM"))
predictor_vars <- predictor_vars[!grepl("^BIO_", predictor_vars)]
print(predictor_vars)
```

---

## Run 100 independent ranger models with `residualEcM` as the response and the phenology predictor set. Each run uses a unique reproducible seed. Per-run summary statistics and OOB predictions are collected into lists and combined into dataframes at the end of the loop.

```{r}
# Set fixed model parameters
ntree      <- 500
ncores     <- 7
n_runs     <- 100
set.seed(22)
base_seeds <- sample.int(1e6, n_runs) # draw n_runs unique seeds for reproducibility

# Initialise storage lists for stats and predictions
run_stats  <- vector("list", n_runs)
pred_list  <- vector("list", n_runs)

for (i in seq_len(n_runs)) {

  seed_i <- base_seeds[i]

  m <- ranger(
    x            = model_df[, ..predictor_vars],
    y            = model_df$residualEcM,
    num.trees    = ntree,
    num.threads  = ncores,
    write.forest = FALSE, # forest object not needed; suppressing saves memory
    seed         = seed_i
  )

  # Collect model-level summary statistics for this run
  run_stats[[i]] <- data.frame(
    run              = i,
    seed             = seed_i,
    r_squared        = m$r.squared,
    prediction_error = m$prediction.error,
    num_trees        = m$num.trees,
    mtry             = m$mtry
  )

  # Collect OOB predictions alongside observed residuals for this run
  pred_list[[i]] <- data.frame(
    run       = i,
    seed      = seed_i,
    cell      = model_df$cell,
    observed  = model_df$residualEcM,
    predicted = m$predictions
  )

  if (i %% 10 == 0) message("Completed run ", i, "/", n_runs) # progress tracker
}

# Combine per-run results into single dataframes
results_stats <- rbindlist(run_stats)
results_preds <- rbindlist(pred_list)
```

---

## Export the per-run model statistics and OOB predictions as CSVs.

```{r}
fwrite(results_stats, file.path(save_dir, "AllResidualModelStats.csv"))
fwrite(results_preds, file.path(save_dir, "AllResidualModelPredictions.csv"))
```
