---
title: "Phenology Contribution — Histogram"
output: html_document
---

This notebook plots the distribution of per-cell delta values, where delta = |combined residual| − |climate residual| averaged across all 100 model runs. A negative delta indicates phenology improved predictions; a positive delta indicates it added noise.

---

## Load packages.

```{r}
library(data.table)
library(ggplot2)
```

---

## Define file paths.

```{r}
model_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
save_dir  <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/PhenologyContributionMap'
```

---

## Load model predictions.

```{r}
climate_preds  <- fread(file.path(model_dir, "AllClimateModelPredictions.csv"))
combined_preds <- fread(file.path(model_dir, "AllCombinedModelPredictions.csv"))
```

---

## Join predictions at the run level, compute per-run delta, and average to cell level.

```{r}
# Join on both cell and run so each row is one run's prediction for one cell
runs <- climate_preds[combined_preds[, .(run, cell, predicted_combined = predicted)],
                      on = .(run, cell), nomatch = 0]

runs[, abs_resid_climate  := abs(observed - predicted)]
runs[, abs_resid_combined := abs(observed - predicted_combined)]
runs[, delta_run          := abs_resid_combined - abs_resid_climate]

# Average per-run delta per cell
cells <- runs[, .(
  observed           = mean(observed),
  predicted_climate  = mean(predicted),
  predicted_combined = mean(predicted_combined),
  delta              = mean(delta_run)
), by = cell]
```

---

## Summary statistics for delta across all GFBI forest cells, including cell-level improvement and worsening rates.

```{r}
n      <- nrow(cells)
se     <- sd(cells$delta) / sqrt(n)
t_crit <- qt(0.975, df = n - 1)
ci_lo  <- mean(cells$delta) - t_crit * se
ci_hi  <- mean(cells$delta) + t_crit * se

delta_summary <- data.frame(
  Statistic = c(
    "Cells where phenology improves predictions (delta < 0)",
    "Cells where phenology worsens predictions (delta > 0)",
    "Mean delta",
    "95% CI (mean)",
    "Median delta",
    "SD delta",
    "Min delta",
    "Max delta"
  ),
  Value = c(
    sprintf("%d (%.1f%%)", sum(cells$delta < 0), mean(cells$delta < 0) * 100),
    sprintf("%d (%.1f%%)", sum(cells$delta > 0), mean(cells$delta > 0) * 100),
    sprintf("%.4f", mean(cells$delta)),
    sprintf("[%.4f, %.4f]", ci_lo, ci_hi),
    sprintf("%.4f", median(cells$delta)),
    sprintf("%.4f", sd(cells$delta)),
    sprintf("%.4f", min(cells$delta)),
    sprintf("%.4f", max(cells$delta))
  )
)

fwrite(delta_summary, file.path(save_dir, "histogram_delta_summary.csv"))

knitr::kable(delta_summary, align = c("l", "r"),
             caption = "Summary statistics for cell-level delta (|combined residual| − |climate residual|) across all GFBI forest cells. 95% CI on the mean computed using the t-distribution.")
```

---

## Plot a histogram of delta values. Green bars indicate cells where phenology improved predictions; red bars where it worsened them.

```{r fig.width=10, fig.height=5}
x_lim <- 0.7

delta_mean   <- mean(cells$delta)
delta_median <- median(cells$delta)
pct_pos      <- mean(cells$delta > 0) * 100
pct_neg      <- mean(cells$delta < 0) * 100

left_label  <- sprintf("%.1f%% cells where phenology improves predictions (green) \n%.1f%% cells where phenology worsens predictions (red)", pct_neg, pct_pos)
right_label <- sprintf(
  "Mean delta = %.4f\n95%% CI = [%.4f, %.4f]\nMedian delta = %.4f\nSD = %.4f\nMin = %.4f\nMax = %.4f",
  delta_mean, ci_lo, ci_hi, delta_median, sd(cells$delta), min(cells$delta), max(cells$delta)
)

p_hist <- ggplot(cells, aes(x = delta, fill = after_stat(x))) +
  geom_histogram(
    binwidth  = 0.01,
    boundary  = 0,
    linewidth = 0.3
  ) +
  geom_rug(sides = "b", alpha = 0.15, linewidth = 0.3, colour = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey30", linewidth = 0.7) +
  scale_fill_gradientn(
    colours = c("darkgreen", "#ccffcc", "#ffcccc", "red"),
    values  = c(0, 0.5 - 1e-6, 0.5 + 1e-6, 1),
    limits  = c(-x_lim, x_lim)
  ) +
  coord_cartesian(xlim = c(-x_lim, x_lim)) +
  annotate("text", x = -x_lim + 0.01, y = Inf, label = left_label,
           hjust = 0, vjust = 2.0, size = 3.5, colour = "black",
           fontface = "plain", lineheight = 1.5) +
  annotate("text", x = x_lim - 0.01, y = Inf, label = right_label,
           hjust = 1, vjust = 1.5, size = 3.5, colour = "black",
           fontface = "plain", lineheight = 1.5) +
  labs(
    x = "Delta (|combined model residual| − |climate model residual|)",
    y = "Number of GFBI Forest Cells"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "none"
  )

p_hist
```



---

## Save the histogram as a PNG.

```{r}
ggsave(file.path(save_dir, "phenology_contribution_histogram.png"),
       plot   = p_hist,
       width  = 10,
       height = 5,
       dpi    = 300)
```

---
