---
title: "Phenology Contribution — Ecoregion Map"
output: html_document
---

This notebook maps the per-cell contribution of phenology aggregated to EPA Level III ecoregions. Delta = |combined residual| − |climate residual| averaged across all 100 model runs and then averaged per ecoregion. A negative mean delta indicates phenology improved predictions on balance within that ecoregion; a positive mean delta indicates it added noise.

---

## Load packages.

```{r}
library(data.table)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(Hmisc)
library(ggspatial)
```

---

## Define file paths.

```{r}
model_dir <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Models'
gee_dir   <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/Data Processing/Google Earth Engine'
save_dir  <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/PhenologyContributionMap'
eco_dir   <- '/Users/williamtucker/Library/CloudStorage/OneDrive-UniversityofEdinburgh/Y4 BIOLOGY (2025-26)/Research Project/Data Analysis/PhenologyContributionMap/NA_CEC_Eco_Level3'
```

---

## Load model predictions and cell coordinates.

```{r}
climate_preds  <- fread(file.path(model_dir, "AllClimateModelPredictions.csv"))
combined_preds <- fread(file.path(model_dir, "AllCombinedModelPredictions.csv"))
coords         <- fread(file.path(gee_dir,   "GFBI_VIIRS_Cells_Data_1.csv"))
```

---

## Join predictions at the run level, compute per-run delta, and average to cell level.

```{r}
# Join on both cell and run so each row is one run's prediction for one cell
runs <- climate_preds[combined_preds[, .(run, cell, predicted_combined = predicted)],
                      on = .(run, cell), nomatch = 0]

runs[, abs_resid_climate  := abs(observed - predicted)]
runs[, abs_resid_combined := abs(observed - predicted_combined)]
runs[, delta_run          := abs_resid_combined - abs_resid_climate]

# Average per-run delta per cell
cells <- runs[, .(
  observed           = mean(observed),
  predicted_climate  = mean(predicted),
  predicted_combined = mean(predicted_combined),
  delta              = mean(delta_run)
), by = cell]
```

---

## Join cell coordinates and build the North America basemap.

```{r}
# Join lon/lat onto the delta table
cells <- coords[, .(cell, lon, lat)][cells, on = "cell", nomatch = 0]

north_america <- ne_countries(
  continent   = "North America",
  scale       = "medium",
  returnclass = "sf"
)
```

---

## Load the EPA Level III North America ecoregion shapefile and reproject to WGS84.

```{r}
# Read shapefile and reproject from Lambert Azimuthal Equal Area to WGS84
ecoregions <- st_read(file.path(eco_dir, "NA_CEC_Eco_Level3.shp"), quiet = TRUE)
ecoregions <- st_transform(ecoregions, crs = 4326)

# Disable s2 spherical geometry and repair any invalid polygons before spatial join
sf::sf_use_s2(FALSE)
ecoregions <- st_make_valid(ecoregions)
```

---

## Spatially join all cells to ecoregions and compute mean delta per ecoregion.

```{r}
# Convert all cell coordinates to an sf points object
cells_sf <- st_as_sf(cells, coords = c("lon", "lat"), crs = 4326)

# Spatial join: assign each cell to the ecoregion polygon it falls within
cells_eco_all <- st_join(cells_sf, ecoregions[, c("NA_L3CODE", "NA_L3NAME")], join = st_within)

# Compute mean delta and 95% CI per ecoregion from all cells
eco_delta_all <- as.data.table(cells_eco_all)[!is.na(NA_L3CODE), .(
  mean_delta = mean(delta),
  ci95       = qt(0.975, df = .N - 1) * sd(delta) / sqrt(.N),
  n_cells    = .N
), by = .(NA_L3CODE, NA_L3NAME)]

cat(sprintf("Ecoregions with cells: %d\n", nrow(eco_delta_all)))
cat(sprintf("  Positive mean delta (phenology hurts on balance): %d\n", sum(eco_delta_all$mean_delta > 0)))
cat(sprintf("  Negative mean delta (phenology helps on balance): %d\n", sum(eco_delta_all$mean_delta < 0)))
cat(sprintf("  Range of mean delta: %.4f to %.4f\n", min(eco_delta_all$mean_delta), max(eco_delta_all$mean_delta)))

eco_delta_all[, ecoregion := paste0(NA_L3CODE, " - ", NA_L3NAME)]
fwrite(eco_delta_all[, .(ecoregion, mean_delta, ci95, n_cells)][order(-mean_delta)],
       file.path(save_dir, "ecoregion_mean_delta.csv"))
```

---

## Summarise ecoregion-level delta: counts of ecoregions with positive and negative mean delta, broken down by whether the 95% CI excludes zero.

```{r}
n_total   <- nrow(eco_delta_all)
n_pos     <- sum(eco_delta_all$mean_delta > 0)
n_pos_sig <- sum(eco_delta_all$mean_delta > 0 & eco_delta_all$mean_delta - eco_delta_all$ci95 > 0)
n_neg     <- sum(eco_delta_all$mean_delta < 0)
n_neg_sig <- sum(eco_delta_all$mean_delta < 0 & eco_delta_all$mean_delta + eco_delta_all$ci95 < 0)

summary_table <- data.frame(
  Direction = c(
    "Positive mean δ (phenology worsens predictions)",
    "Negative mean δ (phenology improves predictions)"
  ),
  `Total ecoregions` = c(n_pos, n_neg),
  `Significant (95% CI excludes zero)` = c(n_pos_sig, n_neg_sig),
  `% Significant` = round(c(n_pos_sig / n_pos, n_neg_sig / n_neg) * 100, 1),
  check.names = FALSE
)

cat(sprintf("Total ecoregions: %d\n", n_total))
cat(sprintf("Overall mean delta across all ecoregions: %.4f\n\n", mean(eco_delta_all$mean_delta)))

fwrite(summary_table, file.path(save_dir, "ecoregion_delta_summary.csv"))

knitr::kable(summary_table, align = c("l", "r", "r", "r"),
             caption = "Summary of ecoregion-level mean delta. Significance is assessed by whether the 95% CI of the mean (t-distribution) excludes zero.")
```

---

## Plot ecoregion choropleth — mean delta per ecoregion across all cells.

```{r fig.width=10.12, fig.height=6.69}
eco_map_all <- merge(ecoregions, eco_delta_all[, .(NA_L3CODE, mean_delta, n_cells)],
                     by = "NA_L3CODE", all.x = TRUE)

p_eco_choropleth_all <- ggplot() +
  geom_sf(data = north_america, fill = "grey85", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = eco_map_all, aes(fill = mean_delta), colour = "grey30", linewidth = 0.2) +
  coord_sf(
    xlim   = c(-180, -40),
    ylim   = c(7, 80),
    expand = FALSE,
    datum  = sf::st_crs(4326)
  ) +
  scale_fill_gradientn(
    name    = "Mean Delta",
    colours = c("darkgreen", "#ccffcc", "#ffcccc", "red"),
    values  = scales::rescale(c(-0.125, -1e-9, 1e-9, 0.125),
                              from = c(-0.125, 0.125)),
    limits  = c(-0.125, 0.125),
    na.value = "grey85"
  ) +
  labs(x = "Longitude", y = "Latitude") +
  annotation_scale(location = "bl", width_hint = 0.25) +
  annotation_north_arrow(
    location = "br", which_north = "true",
    style    = north_arrow_fancy_orienteering()
  ) +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank())

p_eco_choropleth_all
```

---

## Save the ecoregion choropleth as a PNG and PDF.

```{r}
ggsave(file.path(save_dir, "phenology_contribution_map_ecoregions_all.png"),
       plot   = p_eco_choropleth_all,
       width  = 10,
       height = 7,
       dpi    = 300)
```

---

## Plot a violin panel for the 8 ecoregions with a significant negative mean delta (95% CI excludes zero).

```{r fig.width=20, fig.height=13}
shorten_name <- function(x) trimws(gsub(" with .*$", "", x))

# Ecoregions where phenology significantly improves predictions (95% CI wholly negative), most negative first
neg_sig <- eco_delta_all[mean_delta < 0 & mean_delta + ci95 < 0][order(mean_delta)][seq_len(min(8, .N))]

neg_sig[, plot_name := shorten_name(NA_L3NAME)]
neg_sig[, eco_label := factor(
  paste0(NA_L3CODE, " - ", plot_name),
  levels = paste0(NA_L3CODE, " - ", plot_name)[order(mean_delta)]
)]

cells_sub <- as.data.table(cells_eco_all)[NA_L3NAME %in% neg_sig$NA_L3NAME & !is.na(NA_L3NAME)]
cells_sub[neg_sig, eco_label := i.eco_label, on = "NA_L3NAME"]

lbl <- cells_sub[, .(
  n          = .N,
  mean_delta = mean(delta),
  sd_delta   = sd(delta),
  ci95       = qt(0.975, df = .N - 1) * sd(delta) / sqrt(.N),
  y_max      = max(delta),
  y_min      = min(delta)
), by = eco_label]
lbl[, label_top    := sprintf("Mean = %.4f\n95%% CI ± %.4f", mean_delta, ci95)]
lbl[, label_bottom := sprintf("cells = %d\nSD = %.4f", n, sd_delta)]

p_eco_violin <- ggplot(cells_sub, aes(x = eco_label, y = delta)) +
  geom_violin(fill = "#ccffcc", colour = "grey40", linewidth = 0.3, scale = "width", width = 0.7) +
  stat_summary(fun.data = mean_cl_normal, fun.args = list(conf.int = 0.95),
               geom = "errorbar", colour = "darkgreen", width = 0.15, linewidth = 0.6) +
  stat_summary(fun = mean, geom = "point", colour = "darkgreen", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  scale_y_continuous(limits = c(-1.0, 1.0), expand = expansion(0)) +
  geom_text(data = lbl, aes(x = eco_label, y = y_max, label = label_top),
            vjust = -0.8, size = 4.8, lineheight = 1.3, colour = "grey20") +
  geom_text(data = lbl, aes(x = eco_label, y = y_min, label = label_bottom),
            vjust = 1.6, size = 4.8, lineheight = 1.3, colour = "grey20") +
  labs(x = NULL, y = "Delta (|combined residual| − |climate residual|)") +
  theme_bw(base_size = 16) +
  theme(panel.grid.minor  = element_blank(),
        axis.text.x       = element_text(size = 15, angle = 50, hjust = 1, face = "bold"),
        axis.text.y       = element_text(face = "bold"),
        axis.title.y      = element_text(size = 18))

p_eco_violin
```

---

## Save the ecoregion violin plot as a PNG.

```{r}
ggsave(file.path(save_dir, "phenology_contribution_ecoregion_violin.png"),
       plot   = p_eco_violin,
       width  = 20,
       height = 13,
       dpi    = 300)
```

---

```{r}
citation("Hmisc")
```

